title: "Plume flow data"
author: "Nick Bailey"
date: "16/01/2021"
output: html_document
---

# Analysis of data downloaded from Plume Flow2


## Downloaded zip file structure

Set up packages. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages
pacman::p_load(sf, vroom, lubridate, here, tidyverse)

```


Define some key lists, locations, polygons.

```{r locations, include=FALSE}

# pollutants
pollutants <-       c("pm_10", "pm_2.5", "pm_1", 
                      "NO2", "VOC")
pollutants_plume <- c("pm_10_plume", "pm_2.5_plume", "pm_1_plume", 
                      "NO2_plume", "VOC_plume")

# locations - points used in analysis
# - school - jn of Woodcroft and Edgehill
sf_locs <- st_as_sf(data.frame(names =     c(   "home",  "school", "broom_nw", "broom_se"), 
                               latitude =  c(55.880830, 55.879677, 55.881979, 55.875713), 
                               longitude = c(-4.322144, -4.319765, -4.326233, -4.314522)),
                    coords = c("longitude", "latitude"))
st_crs(sf_locs) <- 4326


# buffers round home and school
# - first need to transform to appropriate CRS for st_buffer to work
sf_locs_gb <- st_transform(sf_locs, crs = "EPSG:27700")  # locations with GB grid

# home
sf_home_50m <- st_buffer(sf_locs_gb[1,], dist = 50)
sf_home_50m <- st_transform(sf_home_50m, 4326)   # back to lat/long
# school 150m
sf_school_150m <- st_buffer(sf_locs_gb[2,], dist = 150)
sf_school_150m <- st_transform(sf_school_150m, 4326)   # back to lat/long
# school 300m
sf_school_300m <- st_buffer(sf_locs_gb[2,], dist = 300)
sf_school_300m <- st_transform(sf_school_300m, 4326)   # back to lat/long


# create Broomhill polygon and sf object
broom_poly = st_polygon(list(rbind(c(-4.325908, 55.882598), 
                               c(-4.314163, 55.880051), 
                               c(-4.314145, 55.875598), 
                               c(-4.326992, 55.875881), 
                               c(-4.325908, 55.882598))))
sf_broom = st_sfc(broom_poly, 
                  crs = 4326)
# plot(sf_broom)
```

Identify the most recent .zip file, then get structure and then file names.

Zip file contains:

* 'measures' file (csv) - air quality readings (N=5 in two formats) with date/time
* 'positions' file (csv) - lat/long with date/time
* 'positions' file (kml)
* images directory with png file(s)

```{r basics}

# zip file - name of most recent .zip in 'data' directory
# - doesn't work in one go for some reason
zip_file <- data.frame(names = list.files(here("data"))) 
zip_file <- zip_file %>%
  cbind(file.info(here("data", zip_file$names))) %>%
  filter(grepl("zip", names)) %>%
  filter(mtime == max(mtime)) %>%
  pull(names)

# zip file structure
zip_file_structure <- unzip(here("data", zip_file), list = TRUE) 

# file name - 'measures' csv
file_measures <- zip_file_structure %>%
  filter(grepl("measures", Name)) %>%
  pull(Name)

# file name - 'positions' csv
file_positions_csv <- zip_file_structure %>%
  filter(grepl("positions", Name)) %>%
  filter(grepl("csv", Name)) %>%
  pull(Name)

# file name - 'positions' kml
file_positions_kml <- zip_file_structure %>%
  filter(grepl("positions", Name)) %>%
  filter(grepl("kml", Name)) %>%
  pull(Name)


```

## Read in measures and positions data

Note that the timings for which position is recorded are not the same as the timings for which measures are recorded so will need some approximation to join up. See below.  

Measures file has timestamp and datetime, and then measures for each of the five items recorded in original units and on Plume's standardised scale. 

```{r csv measures}

# read 'measures' csv and tidy colnames
#  NB bit clumsy but works
df_meas <- vroom(map(file_measures, ~ unz(here("data", zip_file), .x))) %>%
  rename(datetime = "date (UTC)") %>%
  rename_with(~ sub(" (Plume AQI)", "_plume", .x, fixed = TRUE)) %>%
  rename_with(~ sub(" (ppb)", "", .x, fixed = TRUE)) %>%
  rename_with(~ sub(" (ug/m3)", "", .x, fixed = TRUE)) %>%
  rename_with(~ sub(" ", "_", .x, fixed = TRUE))

# datetime rounded to nearest whole minute - for joining
df_meas <- df_meas %>% 
  mutate(datetime_hm = round_date(datetime, unit = "minute"))

# number of observations
nrow(df_meas)

```


Positions from csv. Has lat/long but also timestamp and datetime (POSIXct). We need this as datetime not in the kml file. 

```{r positions csv}

# read 'positions' csv
df_pos_csv <- vroom(map(file_positions_csv, ~ unz(here("data", zip_file), .x))) %>% 
  rename(datetime = date)

# summary
str(df_pos_csv)


```


Positions from kml. Has only two fields (Name and Description) both of which appear to be empty so attach datetime from csv. 

```{r positions kml}

# read 'positions' kml
sf_pos_kml <- read_sf(unzip(here("data", zip_file), files = file_positions_kml)) %>% 
  cbind(df_pos_csv$datetime) %>% 
  rename(datetime = df_pos_csv.datetime) 

# datetime rounded to nearest whole minute - for joining
sf_pos_kml <- sf_pos_kml %>% 
  mutate(datetime_hm = round_date(datetime, unit = "minute"))

# tidy up - delete folders+files created by unzip
unlink("flow", recursive = TRUE)

# quick plot
plot(sf_pos_kml[1])   # to limit to one plot

```

Map of all data points with Broomhill boundary, and home + school marked. 

``` {r map all}

ggplot(data = sf_pos_kml) +
  geom_sf(data = sf_broom, fill = NA) +
  geom_sf(data = sf_locs[1:2,], size = 4, shape = 23, fill = "darkred") +
  geom_sf(size = .5, alpha = .5) 
    

```


Zoomed in to Broomhill neighbourhood with very rough Broomhill boundary. 

``` {r map broomhill}

sf_pos_kml %>% 
  ggplot() +
    geom_sf(data = sf_broom, fill = NA) +
    geom_sf(data = sf_locs[1:2,], size = 4, shape = 23, fill = "darkred") +
    geom_sf(fill = "blue", size = 2, alpha = 0.5)  +
    coord_sf(xlim = c(-4.33, -4.31), ylim = c(55.875, 55.884), expand = FALSE)
    
```

## Dates and times


Days for which we have positional data. 

``` {r fig days}

sf_pos_kml %>% 
  ggplot(aes(x = wday(datetime, week_start = 1, label = TRUE))) +
  geom_bar() 
  

```
Hours of the day for which we have positional data.

``` {r fig hours}

sf_pos_kml %>% 
  ggplot(aes(x = hour(datetime))) +
  geom_histogram(bins = 24)
  

```

Minutes within the hour. 

``` {r fig mins}

sf_pos_kml %>% 
  ggplot(aes(x = minute(datetime))) +
  geom_histogram(bins = 60)
  
```
And seconds. 

``` {r fig secs}

sf_pos_kml %>% 
  ggplot(aes(x = second(datetime))) +
  geom_histogram(bins = 60)
  

```

Intervals between measure - (almost) always one minute.  

``` {r fig interval meas}

df_meas %>% 
  mutate(interval_mins = round((datetime - lag(datetime))/60,0)) %>% 
  ggplot(aes(x = interval_mins)) +
  geom_bar() +
  coord_cartesian(xlim = c(-100, 100))
  

```

Intervals between positions in minutes. The large majority are one minute or less. 

NB that difference between two times has class 'difftime' (measured in seconds but here converted to minutes). 

``` {r fig interval pos}

sf_pos_kml %>% 
  mutate(interval_mins = round((datetime - lag(datetime))/60,0)) %>% 
  ggplot(aes(x = interval_mins)) +
  geom_bar() + 
  coord_cartesian(xlim = c(0, 20))
  

```

Number of position observations in the same minute. In most cases, there is more than one position record in each minute. So we lose a lot of positional date by aggregating to minutes - but then we only have measurements once per minute. 

``` {r pos per min}

df_pos_csv %>%
  mutate(datetime_hm = round_date(datetime, unit = "minute")) %>% 
  group_by(datetime_hm) %>% 
  summarise(N = n()) %>%
  ggplot(aes(x = N)) +
  geom_histogram()

```

## Pollution data

Boxplot of pollutants shows some extreme values, esp. PM 1 and VOCs. 

``` {r pollutant}

df_meas %>% 
  pivot_longer(all_of(pollutants), names_to = "pollutant", values_to = "level") %>% 
  ggplot(aes(y = level)) +
  facet_wrap(~ pollutant, scales = "free") +
  geom_boxplot()
  

```

``` {r pollutant plume}

df_meas %>% 
  pivot_longer(all_of(pollutants_plume), names_to = "pollutant", values_to = "level") %>% 
  ggplot(aes(x = pollutant, y = level)) +
  geom_boxplot()
  

```
Look at spread over time. there are a couple of real outlier days where some readings off the scale. 

``` {r pollutant time, fig.height = 8, fig.width = 8}

df_meas %>% 
  pivot_longer(all_of(pollutants), names_to = "pollutant", values_to = "level") %>% 
  mutate(date = round_date(datetime, unit = "hour")) %>% 
  ggplot(aes(x = date, y = level, fill = pollutant)) +
  geom_point(aes(colour = pollutant)) +
  facet_wrap(~ pollutant, ncol = 1, scale = "free")
  

```

Look at spread over time - plume version.

``` {r pollutant time, fig.height = 8, fig.width = 8}

df_meas %>% 
  pivot_longer(all_of(pollutants_plume), names_to = "pollutant", values_to = "level") %>% 
  mutate(date = round_date(datetime, unit = "hour")) %>% 
  ggplot(aes(x = date, y = level, fill = pollutant)) +
  geom_point(aes(colour = pollutant)) +
  facet_wrap(~ pollutant, ncol = 1, scale = "free")
  

```

## Joining pollution measures to position
Easiest way to join is by rounding datetime in both position and measure databases to the nearest minute, and attaching the relevant measure record based on the same rounding. There will be multiple positions records with the same time in rounded minutes, each matched to the same measures record (since these are every minute). We therefore select the positions record which is nearest in timing to the relevant measures record (both measured to the second). 

Note that this isn't necessarily the closest record in time since the rounding process could put that in a different group (by rounded minustes). And a more sophisticated approach would be to attach the measures records for timings either side of position timing, and interpolate a measure for the specific timing of the position. But that seems more work than necessary at the moment. 

We can then make maps of pollution - here, PM10. 

``` {r join pos meas}

sf_pos_meas_kml <- sf_pos_kml %>% 
  left_join(df_meas, by = "datetime_hm") %>%
  mutate(datetime_diff = abs(datetime.x - datetime.y)) %>%    #abs value of diff
  group_by(datetime_hm) %>% 
  filter(datetime_diff == min(datetime_diff))

sf_pos_meas_kml %>% 
  ggplot() +
  geom_sf(data = sf_broom, colour = "darkred", fill = NA) +
  geom_sf(aes(colour = pm_10), size = 3) +
  scale_colour_continuous(type = "viridis", direction = -1) +
  geom_sf(data = sf_locs[1:2,], size = 3, shape = 23, fill = "darkred") 

```

## Points in Broomhill and around Broomhill School

Identify points within school buffer and plot. 

``` {r broomhill}

# list of points in Broomhill or within school buffers 
df_broom <- as.data.frame(st_within(sf_pos_meas_kml, sf_broom, sparse = FALSE)) %>%
  rename(yn = V1)
df_home_50m <- as.data.frame(st_within(sf_pos_meas_kml, sf_home_50m, sparse = FALSE)) %>%
  rename(yn = V1)
df_school_150m <- as.data.frame(st_within(sf_pos_meas_kml, sf_school_150m, sparse = FALSE)) %>%
  rename(yn = V1)
df_school_300m <- as.data.frame(st_within(sf_pos_meas_kml, sf_school_300m, sparse = FALSE)) %>%
  rename(yn = V1)


# map of points in Broomhill
sf_pos_meas_kml %>% 
  cbind(df_broom) %>%
  filter(yn == 1) %>% 
  ggplot() +
  geom_sf(data = sf_broom, colour = "darkred", fill = NA) +
  geom_sf(data = sf_home_50m, colour = "darkred", fill = NA) +
  geom_sf(data = sf_school_300m, colour = "darkred", fill = NA) +
  geom_sf(aes(colour = pm_10), size = 3) +
  scale_colour_continuous(type = "viridis", direction = -1) +
  geom_sf(data = sf_locs[1:2,], size = 3, shape = 23, fill = "darkred") +
    coord_sf(xlim = c(-4.33, -4.31), ylim = c(55.875, 55.884), expand = FALSE)

```

``` {r broomhill}

# map of points in 300m of school but not home
sf_pos_meas_kml %>% 
  cbind(df_school_300m) %>%
  rename(yn_school = yn) %>% 
  cbind(df_home_50m) %>%
  rename(yn_home = yn) %>% 
  filter(yn_school == 1 & yn_home == 0) %>% 
  ggplot() +
  geom_sf(data = sf_broom, colour = "darkred", fill = NA) +
  geom_sf(data = sf_school_300m, colour = "darkred", fill = NA) +
  geom_sf(aes(colour = pm_10), size = 3) +
  scale_colour_continuous(type = "viridis", direction = -1) +
  geom_sf(data = sf_locs[1:2,], size = 3, shape = 23, fill = "darkred") +
    coord_sf(xlim = c(-4.33, -4.31), ylim = c(55.875, 55.884), expand = FALSE)

```

For points within school buffer but not home buffer, pollutants by time of day. 

``` {r school time}

# scatter points close to school by time of day
sf_pos_meas_kml %>% 
  cbind(df_school_300m) %>%
  rename(yn_school = yn) %>% 
  cbind(df_home_50m) %>%
  rename(yn_home = yn) %>% 
  filter(yn_school == 1 & yn_home == 0) %>% 
  mutate(hours = hour(datetime.x) + minute(round_date(datetime.x, unit = "minute"))/60) %>%
  filter(hours > 8.5 & hours < 9.5) %>% 
  pivot_longer(all_of(pollutants_plume), names_to = "pollutant", values_to = "level") %>% 
  ggplot(aes(x = hours, y = level)) +
  geom_point(aes(colour = pollutant)) +
  geom_smooth(aes(colour = pollutant), se = FALSE) 
  
```


For points within school buffer but not home buffer, pollutants by time of day - facets. 

``` {r school time}

# scatter points close to school by time of day
as.data.frame(sf_pos_meas_kml) %>% 
  cbind(df_school_300m) %>%
  rename(yn_school = yn) %>% 
  cbind(df_home_50m) %>%
  rename(yn_home = yn) %>% 
  filter(yn_school == 1 & yn_home == 0) %>% 
  mutate(hours = hour(datetime.x) + minute(round_date(datetime.x, unit = "minute"))/60) %>%
  filter(hours > 8.5 & hours < 9.5) %>% 
  pivot_longer(all_of(pollutants), names_to = "pollutant", values_to = "level") %>% 
  ggplot(aes(x = hours, y = level)) +
  geom_point(aes(colour = pollutant)) +
  facet_wrap(~ pollutant, scales = "free") +
  geom_smooth(aes(colour = pollutant), se = FALSE) 
  
```

For points within school buffer but not home buffer, pollutants (Plume) by time of day - facets. 

``` {r school time}

# scatter points close to school by time of day
as.data.frame(sf_pos_meas_kml) %>% 
  cbind(df_school_300m) %>%
  rename(yn_school = yn) %>% 
  cbind(df_home_50m) %>%
  rename(yn_home = yn) %>% 
  filter(yn_school == 1 & yn_home == 0) %>% 
  mutate(hours = hour(datetime.x) + minute(round_date(datetime.x, unit = "minute"))/60) %>%
  filter(hours > 8.5 & hours < 9.5) %>% 
  pivot_longer(all_of(pollutants_plume), names_to = "pollutant", values_to = "level") %>% 
  ggplot(aes(x = hours, y = level)) +
  geom_point(aes(colour = pollutant)) +
  facet_wrap(~ pollutant, scales = "free") +
  geom_smooth(aes(colour = pollutant), se = FALSE) 
  
```